Point transect distance sampling in inlabru

### 2019-08-09 paper.tex
Update paper.tex
Added shortcuts, Case Study text and links to figures.

% shortcuts
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\T}{\intercal}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bm}{\boldsymbol}  % bold maths symbols
\newcommand{\tl}{\tilde{\lambda}}   % thinned little lambda
\newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author{Andrew E Seaton}

\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}

@@

@@

>>>>

% shortcuts
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\T}{\intercal}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bm}{\boldsymbol}  % bold maths symbols
\newcommand{\tl}{\tilde{\lambda}}   % thinned little lambda
\newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda
% RJC 09/08/2019 Added shortcuts for Hawaiian words
\newcommand{\akepa}{\textquotesingle\={a}kepa}  % adds Hawaiian diacritical marks
\newcommand{\Akepa}{\textquotesingle\={A}kepa}  % adds Hawaiian diacritical marks
\newcommand{\hawaii}{Hawai\textquotesingle i}   % adds Hawaiian diacritical marks

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label,2]{<address>}

% RJC 09/08/2019 Added co-authors and affiliations. Note that I used \affil[]{} instead of \address[]{}
\author[1,*]{Andrew E Seaton}
\author[1,2]{Richard J Camp}
\author[1]{Janine B Illian}
\author[1]{David L Borchers}
\author[1]{David L Miller}      % Ask about co-authoring the manuscript
\author[1]{Len Thomas}          % Ask about co-authoring the manuscript
\author[1]{Stephen T Buckland}  % Ask about co-authoring the manuscript
\author[3]{Steve J Kendall}     % It is Steve's data we are using.

%\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
\affil[1]{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
\affil[2]{U. S. Geological Survey, Pacific Island Ecosystems Research Center, P.O. Box 44, \hawaii{} National Park, HI 96718, U.S.A.}
\affil[3]{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
\affil[*]{Correspondence: Andrew E Seaton, Email: aes22@st-andrews.ac.uk}

@@

% RJC 09/08/2019 Added text describing the akepa data, survey method, key questions and what we care about from ecology/conservation point of view. This text provides a background on Hawaiian forest birds, why it is important to monitor their populations and the need for unbiased and precise estimates. I am wondering if we need to strengthen the 'key questions' text.

@@

% RJC 09/08/2019 Added case study text between % addtextstart and % addtextend markers.
% addtextstart
In this section, we present a description of Hawaiian forest birds, forest bird monitoring, the study species, study area, bird sampling, and description of the data. \hawaii{} is home to the largest adaptive radiation of oceanic archipelago birds worldwide. The Hawaiian Honeycreepers (Drepanidinae) is comprised of 21 genera and more than 50 species generally grouped according to feeding niches with many species and subspecies being island endemics. The honeycreepers, like most insular bird species, evolved life history strategies that include greater longevity with high adult survival, delayed maturation with an extended post-fledging dependency period, decreased clutch size and fecundity, and reduced dispersal abilities with high site fidelity \citep{woodworth_demography_2009}. While the honeycreepers diversified into the plethora of ecological niches they also had a loss or reduction of defences against competitors, predators and diseases. The loss of defensive adaptations made the honeycreepers particularly vulnerable to humans and the changes and threats they brought. As a result, more than half of the honeycreepers are extinct \citep{banko_extinctions_2009}. A mere 23 honeycreeper species persist with most species restricted to minute portions of their original ranges and in small numbers. The IUCN lists 21 honeycreeper species as critically endangered, endangered or vulnerable, and only two species are categorized as least concern \citep{iucn_redlist_2018}.

In the late $19^{\textnormal{th}}$ century naturalists to \hawaii{} described the Hawaiian forest birds and documented species distributions and abundances. Subsequent naturalists to the islands documented range contractions and population declines of native species while noting expansion in non-native bird species ranges and abundances. It was not until standardized and statistically robust sampling techniques to estimate population sizes and associated standard errors were initiated that the extent to which honeycreepers had declined was appreciated. Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}. Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species. Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.

\subsection*{Study species}
The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA. \Akepa{} is a foraging specialist whose diet consists of spiders, insects and psyllids, as well as nectar. \Akepa{} uses its cross-bill to probe and open terminal leaf clusters and extracts prey with its tongue. During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}. \Akepa{} has a global abundance of 16,248 (95\%CI 10,074\textendash25,198) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}. Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau; 19$^{\circ}$ 51\textquotesingle N, 155$^{\circ}$ 18\textquotesingle W), supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}. Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern. Moreover, having unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.

\subsection*{Study area}
Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants. Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance. Survey points were established along 15 transects following a systematic, random design with points approximately 150-m apart on transects located either 500 or 1,000 m apart. Following \cite{camp_population_2010, camp_statespace_2016}, we restrict our study area to the 3,061 ha open-forest stratum of Hakalau. The study area is between 1,400 and 1,920 m elevation of montane forest dominated by native \textquotesingle \={o}hi\textquotesingle a (\textit{Metrosideros polymorpha}) and koa (\textit{Acacia koa}) with a mixture of native and non-native understory plants. The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}. To the north the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig. \ref{fig:2002studyareapointspt}). The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of their study area. The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa. The substrate in the southern portion of the study area is nutrient-rich, old lava flows (11,000 to 64,000 yrs old) whereas the northern portion occurs on new lava flows (5,000 to 10,000 yrs old) that are nutrient depleted \citep{ccp_hakalau_2010}.

\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{figures/2002studyareapoints_pt}
	\caption{Study area (light blue) showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.}
	\label{fig:2002studyareapointspt}
\end{figure}

\subsection*{Bird sampling}
Surveys used point-transect methods, a form of distance sampling, where horizontal distances from survey points to individually detected birds are used to model a species-specific detection probability that is used to estimate absolute abundance and variance \citep{buckland_distance_2015}. Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength \textgreater Baufort scale 3). During 8-min counts trained observers recorded the species, exact distance to the nearest meter and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.

\subsection*{Data description}
For purposes of our analysis, we selected a single survey from the \akepa{} time series based on broad sampling of the study area and sufficient numbers of detections. In 2002, 195 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig. \ref{fig:2002studyareapointspt}). On 67 points 152 \akepa{} were detected. The number of detections ranged from zero to 6 (Fig. \ref{fig:2002countspt}). These data are used for both the two-stage and one-stage model-based analyses. \cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the open-forest study area and the bird surveys.

\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{figures/2002counts_pt}
	\caption{Counts of \akepa{} by point during the 2002 survey.}
	\label{fig:2002countspt}
\end{figure}

% addtextend
